A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-13 jour/jour.tex
more

We fix a signature $\Sig = (\Types, \Funcs)$. The types are partitioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
\emph{datatypes}, $\Codata$ are the \emph{codatatypes}, and $\Nondata$ are the remaining
\emph{ordinary types} (which can be interpreted or not). The function symbols are
partitioned into $\Funcs = \Ctr \mathrel{\uplus} \Sel$, where $\Ctr$ are the
\emph{constructors} and $\Sel$ are the \emph{selectors}. There is no need to
consider further function symbols because they can be abstracted away as
variables when combining theories.
Exceptionally, it is convenient to use numeric constants ($0$, $1$, \ldots)\ in
examples.
$\mathrm\Sigma$-terms are standard first-order terms
over $\mathrm\Sigma$,
without $\mu$-binders.
%(Section~\ref{sec:implementation-as-a-theory-solver-in-cvc4}).

@@

The decision procedure is implemented as a \relax{theory solver} of CVC4---that is, a
specialized solver for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the CDCL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\: a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of
\begin{itemize}
\item the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
\item additional (dis)equalities between variables of types not belonging to
$T_i$.
\end{itemize}
Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\: C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

\item Repeat until $\mathcalx{U}$ is unchanged:
\begin{enumerate}
\item[2.1.] $\mathcalx{U}' := \emptyset$.
\item[2.2.] For each $U_i \in \mathcalx{U}$:
\begin{itemize}
\item[2.2.1.] Let $\mathcalx{U}_i$ be a partition of $U_i$ such that $\ec{t} =_{\mathcalx{U}_i} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$, and
\item $\ec{t_j} =_{\mathcalx{U}} \ec{u_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[2.2.2.] $\mathcalx{U}' := \mathcalx{U}' \cup \mathcalx{U}_i$.
\end{itemize}
\item[2.3.] $\mathcalx{U} := \mathcalx{U}'$.
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcalx{U}$, $\rn{Unique}(t,u)$.
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Unique}}
\label{fig:uniqueness-algo}
\end{figure}

@@

For \rn{Unique}, Fig.~\ref{fig:uniqueness-algo} considers a partition $\mathcalx{U}$ of the equivalence classes of our set $\Ec$
such that codatatype equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, 
for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcalx{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcalx{U}$.
We then refine $\mathcalx{U}$ by 
constructing a partition $\mathcalx{U}_i$ of each $U_i \in \mathcalx{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcalx{U}_i$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcalx{U}$ for each $j=1, \ldots, k$.
Let $\mathcalx{U'}$ be the union of each of these partitions.
We update $\mathcalx{U}$ to $\mathcalx{U'}$.
This refinement is repeated until $\mathcalx{U}$ is left unchanged.
If the resulting partition $\mathcalx{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.
This algorithm is analogous to Hopcroft's algorithm for minimizing
deterministic finite automata \cite{hopcroft-1971}---terms correspond to
states, argument positions $1,\ldots,k$ correspond to input symbols, and the top constructor symbol
of a term generalizes the accepting/rejecting status of a state.

@@

When instructed to do so, the implementation produces models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we %may
construct a map $\ValC$ from 
the equivalence classes of $\Ec$ to closed normal $\MU$-terms.
Recall that our construction of $\ValC$ requires picking values $\MU {\Varec{x}}.\; t$ for each $\Varec{x} \in \FV( \ValC )$
such that $\MU {\Varec{x}}.\; t$ does not occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To pick values $\MU {\Varec{x}}.\; t$, we use a fair enumerator for each (co)datatype $\tau$,
which lists all the values
of type $\tau$ in a normal interpretation. %, which are closed normal $\MU$-terms.

%Concretely,
For codatatypes, we enumerate a stream of \relax{every} $\MU$-term of our signature $\Sigma$ in a fair manner,
discarding those that have free variables or are not normal.
For those that are closed and normal, we then check whether they occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
This set does not need to be explicitly constructed.
%We do not explicitly construct the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To determine if a term $\MU {\Varec{x}}.\; t$ occurs in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$, 
we check if it is $\alpha$-equivalent to a closed term $(\MU {\Varec{y}}.\; u) \{ \Varec{x} \mapsto z \}$ for
some $\MU {\Varec{y}}.\; u$ in the range of $\ValC$ and variable $z$.
This can be efficiently achieved by matching the term $\MU {\Varec{x}}.\; t$ with %the term
$\MU {\Varec{y}}.\; u$.

@@

While the decision procedure is restricted to universal conjectures, users often
want to solve problems that feature universal axioms and existential
conjectures.
%%% @ANDY: not sure how to phrase this now
%where checking satisfiability requires reasoning about ground literals in the
%presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete 
instantiation-based 
techniques~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These techniques extend naturally to %quantified
formulas involving datatypes and codatatypes.

However, the presence of quantifiers poses an additional challenge in
the context of (co)datatypes. Quantified formulas may entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
When assuming that $\ty{u}$ has infinite cardinality, the calculus
presented in Section~\ref{sec:the-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete for quantified formulas.
Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of
uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the $\ty{stream}_{\vthinspace\ty{u}}$ codatatype %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$,
which states that either the cardinality of $\ty{u}$ is greater than one or
$s$ must be equal to $t$.

@@

The decision procedure for (co)datatypes is useful both for proving (via
negation, in the refutational style) and finite model finding.
It is in fact vital for finite model finding,
because the acyclicity and uniqueness rules are necessary for solution
soundness, without which the generated models would often be
spurious. For example, given the axioms
\begin{align*}
\const{zeros} & = \const{SCons}(0,\:\const{zeros})
& \const{sconst}~n & = \const{SCons}(n,\:\const{sconst}~n)
\end{align*}
the conjecture
$\const{zeros} = \const{sconst}~0$
would be ``refuted'' by a spurious countermodel where
\const{zero} and $\const{sconst}~0$ are interpreted by two distinct
values that both correspond to the \hbox{$\mu$-term}
$\MU s.\; \const{SCons}(0, s)$,
violating uniqueness.

>>>>

We fix a signature $\Sig = (\Types, \Funcs)$. The types are partitioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
\emph{datatypes}, $\Codata$ are the \emph{codatatypes}, and $\Nondata$ are the remaining
\emph{ordinary types} (which can be interpreted or not). The function symbols are
partitioned into $\Funcs = \Ctr \mathrel{\uplus} \Sel$, where $\Ctr$ are the
\emph{constructors} and $\Sel$ are the \emph{selectors}. There is no need to
consider further function symbols because they can be abstracted away as
variables when combining theories.
Exceptionally, it is convenient to use numeric constants ($0$, $1$, \ldots)\ in
examples.
$\Sig$-terms are standard first-order terms
over $\Sig$,
without $\mu$-binders.
%(Section~\ref{sec:implementation-as-a-theory-solver-in-cvc4}).

@@

The decision procedure is implemented as a \relax{theory solver} of CVC4---that is, a
specialized procedure for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the CDCL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of~$F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\: a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant \cite{jovanovic2011sharing}
of the Nelson--Oppen method \cite{nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of
\begin{itemize}
\item the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
\item additional (dis)equalities between variables of types not belonging to
$T_i$.
\end{itemize}
Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\: C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

\item Repeat until $\mathcalx{U}$ is unchanged:
\begin{enumerate}
\item[2.1.] $\mathcalx{U}' := \emptyset$.
\item[2.2.] For each $U_i \in \mathcalx{U}$:
\begin{itemize}
\item[2.2.1.] Let $\mathcalx{U}_{\vthinspace i}$ be a partition of $U_i$ such that $\ec{t} =_{\mathcalx{U}_{\vthinspace i}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$, and
\item $\ec{t_j} =_{\mathcalx{U}} \ec{u_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[2.2.2.] $\mathcalx{U}' := \mathcalx{U}' \cup \mathcalx{U}_{\vthinspace i}$.
\end{itemize}
\item[2.3.] $\mathcalx{U} := \mathcalx{U}'$.
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcalx{U}$, $\rn{Unique}(t,u)$.
\end{enumerate}
\vspace*{-\topsep}
\caption{\,Algorithm for checking \rn{Unique}}
\label{fig:uniqueness-algo}
\end{figure}

@@

For \rn{Unique}, Fig.~\ref{fig:uniqueness-algo} considers a partition $\mathcalx{U}$ of the equivalence classes of our set $\Ec$
such that codatatype equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, 
for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcalx{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcalx{U}$.
We then refine $\mathcalx{U}$ by 
constructing a partition $\mathcalx{U}_{\vthinspace i}$ of each $U_i \in \mathcalx{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcalx{U}_{\vthinspace i}$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcalx{U}$ for each $j=1, \ldots, k$.
Let $\mathcalx{U'}$ be the union of each of these partitions.
We update $\mathcalx{U}$ to $\mathcalx{U'}$.
This refinement is repeated until $\mathcalx{U}$ is left unchanged.
If the resulting partition $\mathcalx{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.
This algorithm is analogous to Hopcroft's algorithm for minimizing
deterministic finite automata \cite{hopcroft-1971}---terms correspond to
states, argument positions $1,\ldots,k$ correspond to input symbols, and the top constructor symbol
of a term generalizes the accepting/rejecting status of a state.

@@

When instructed to do so, the implementation produces models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we %may
construct a map $\ValC$ from 
the equivalence classes of $\Ec$ to closed normal $\MU$-terms.
Recall that our construction of $\ValC$ requires choosing values $\MU {\Varec{x}}.\; t$ for each $\Varec{x} \in \FV( \ValC )$
such that $\MU {\Varec{x}}.\; t$ does not occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To choose values, % $\MU {\Varec{x}}.\; t$,
we use a fair enumerator for each (co)datatype $\tau$,
which lists all the values
of type $\tau$ in a normal interpretation. %, which are closed normal $\MU$-terms.

%Concretely,
For codatatypes, we enumerate a stream of \relax{every} $\MU$-term of our signature $\Sig$ in a fair manner,
discarding those that have free variables or are not normal.
For those that are closed and normal, we then check whether they occur in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
This set does not need to be explicitly constructed.
%We do not explicitly construct the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$.
To determine if a term $\MU {\Varec{x}}.\; t$ occurs in the set $\BAD_{\smash{\J}}^{\Varec{x}}( \ValC )$, 
we check if it is $\alpha$-equivalent to a closed term $(\MU {\Varec{y}}.\; u) \{ \Varec{x} \mapsto z \}$ for
some $\MU {\Varec{y}}.\; u$ in the range of $\ValC$ and variable $z$.
This can be efficiently achieved by matching the term $\MU {\Varec{x}}.\; t$ with %the term
$\MU {\Varec{y}}.\; u$.

@@

While the decision procedure is restricted to universal conjectures, users often
want to solve problems that feature universal axioms and existential
conjectures.
%%% @ANDY: not sure how to phrase this now
%where checking satisfiability requires reasoning about ground literals in the
%presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete 
instantiation-based 
techniques~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These techniques extend naturally to quantified
formulas involving datatypes and codatatypes.

However, the presence of quantifiers poses an additional challenge in
the context of (co)datatypes. Quantified formulas can entail an upper bound on
the cardinality of an uninterpreted type $\ty{u}$.
By assuming that uninterpreted types have infinite cardinality, the calculus
presented in Section~\ref{sec:the-decision-procedure} is \relax{incomplete}
since it may fail to recognize cases where \rn{Split} and \rn{Single} should be
applied.
This does not %necessarily
impact the correctness of the procedure in this setting,
since %by assumption
the solver is already incomplete for quantified formulas.

Nonetheless, two techniques help increase the precision of the solver.
First, we can apply \rn{Split} to datatype terms whose cardinality depends on the finiteness of
uninterpreted types.
Second, we can conditionally apply \rn{Single} to codatatype terms that may have cardinality one.
For example, the codatatype $\ty{stream}_{\vthinspace\ty{u}}$ %, where $\ty{u}$ is uninterpreted.
has cardinality one precisely when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s}$ and $\ec{t}$ for this type,
the implementation adds the clause %of the form
$(\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y ) \mathrel\vee s \teq t$ to $F\!$,
which states that either the cardinality of $\ty{u}$ is greater than one or
$s$ must be equal to $t$.

@@

The decision procedure for (co)datatypes is useful both for proving (via
negation, in the refutational style) and for finite model finding.
It is in fact vital for finite model finding,
because the acyclicity and uniqueness rules are necessary for solution
soundness, without which the generated models would often be
spurious. For example, given the axioms
\begin{align*}
\const{zeros} & = \const{SCons}(0,\:\const{zeros})
& \const{sconst}~n & = \const{SCons}(n,\:\const{sconst}~n)
\end{align*}
the conjecture
$\const{zeros} = \const{sconst}~0$
would be ``refuted'' by a spurious countermodel where
\const{zero} and $\const{sconst}~0$ are interpreted by two distinct
values that both correspond to the \hbox{$\mu$-term}
$\MU s.\; \const{SCons}(0, s)$,
violating uniqueness.
